---
title: "US_Accidents_Analysis"
author: "Team: Bhoomika Nanjaraja, Khush Shah, Apoorva Reddy Bagepalli, Devarsh Apurva Sheth"
output:
  html_document:
    code_folding: show
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---
```{r init, include=F}
# The package "ezids" (EZ Intro to Data Science) includes some helper functions we developed for the course. 
# Some of the frequently used functions are loadPkg(), xkabledply(), xkablesummary(), uzscale(), etc.
# You will need to install it (once) from GitHub.
# library(devtools)
# devtools::install_github("physicsland/ezids")
# Then load the package in your R session.
library(ezids)
library(ggplot2)
library(dplyr)
library(lubridate)
library(tidyverse)
library(scales)
library(plotly)
library(gridExtra)
library(tidytext)
library(modelr)
library(caret)
library(ROSE)
library(glmnet)
library(randomForest)
library(ggrepel)
library(patchwork)
library(readr)
```

```{r setup, include=FALSE}
# Some of common RMD options (and the defaults) are: 
# include=T, eval=T, echo=T, results='hide'/'asis'/'markup',..., collapse=F, warning=T, message=T, error=T, cache=T, fig.width=6, fig.height=4, fig.dim=c(6,4) #inches, fig.align='left'/'center','right', 
knitr::opts_chunk$set(warning = F, message = F)
# Can globally set option for number display format.
options(scientific=T, digits = 3) 
# options(scipen=9, digits = 3) 
```
#Introduction

#Data Description

#SMART Questions


# Exploratory Data Analysis

## Reading the US Accidents Related data from CSV File.

```{r}
accidents <- read_csv("/Users/bhoomikan/Documents/Sample_Projects_DS/US_Accidents_EDA.csv")
```

## Map For Accident Distribution In USA

```{r}
address <- c("Country", "City", "County", "Street", "Zipcode")
df_add <- accidents %>% select(-all_of(address))

df_add <- df_add %>% 
  mutate(Severity = as.character(Severity)) %>% 
  mutate_if(is.logical, as.character)

df_add <- df_add %>% 
  type_convert() %>%
  mutate(Severity = factor(Severity)) %>%
  mutate_if(is.logical, factor) %>%
  mutate_if(is.character, factor)

states <- map_data("state") %>% as_tibble() %>% select(long, lat, group, region)
states_abb <- read_csv("/Users/bhoomikan/Documents/Sample_Projects_DS/data.csv") %>%
  mutate(State = tolower(State)) %>%
  select(State, Code) %>%
  rename("State_full" = State)
accident_count <- df_add %>%
  count(State) %>%
  left_join(states_abb, by = c("State" = "Code"))

states <- states %>%
  left_join(accident_count, by = c("region" = "State_full"))
# top 10 states
top_10 <- accident_count %>%
  arrange(desc(n)) %>%
  head(10)
top_10 <- top_10$State %>% unlist()

top_10_map <- states %>%
  filter(State %in% top_10)
top_10_label <- top_10_map %>%
  group_by(region, State) %>%
  summarise(long = mean(long), lat = mean(lat))

ggplot(states, aes(long, lat, group = group)) +
  geom_polygon(aes(fill = n), color = "#636363", size = 0.1) +
  geom_polygon(data = top_10_map, color = "red", fill = NA, size = 0.8) +
  scale_fill_gradient(low = "#fee5d9", high = "#de2d26",
                      name = "Accident Count", labels = unit_format(unit = "K", scale = 1e-03)) +
  ggrepel::geom_label_repel(mapping = aes(label = State, group = 1), data = top_10_label) +
  theme_minimal() +
  coord_quickmap() +
  labs(title = "Accident distribution in the U.S.",
       x = "Longitude",
       y = "Latitude")
```

<b>The map showcases the top 10 states with the highest accident frequencies recorded between 2016 and 2023. Notably, California leads the list with nearly 1.5 million accidents, and Florida follows closely with 829.646 million accidents during this period.</b>

## Bar Graph For Accident Distribution In USA

```{r}
states_abb <- accidents %>% mutate(State = tolower(State)) %>% select(State) %>% rename("State_full" = State)
accident_count <- accidents %>% count(State)

# top 10 states
top_10 <- accident_count %>%
  arrange(desc(n)) %>%
  head(10)
top_10 <- top_10$State %>% unlist()

accidents %>% 
  filter(State %in% top_10) %>%
  count(State) %>%
  ggplot(aes(reorder(State, n), n)) +
  geom_col() +
  geom_label(aes(label = n), nudge_y = -30000) +
  labs(x = NULL, y = "Number of accidents",
       title = "Top 10 States with the most accidents") +
  scale_x_discrete(labels = rev(c("California", "Florida","Texas", "South Carolina","New York",
                                  "North Carolina", "Pennsylvania","Virginia",
                                  "Minnesota", "Oregon"))) +
  scale_y_continuous(breaks = seq(0, 700000, 100000), labels = scales::unit_format(unit = "K", scale = 1e-03)) +
  coord_flip()
```

<b>The bar graph above illustrates the top 10 states with the highest accident frequencies between 2016 and 2023. California stands out as the state with the highest number of accidents during this period, accounting for approximately 20% of the total accidents, with Florida following closely.</b>

## Comparison of proportion of accidents occuring at different times of the day

```{r}
LATE_NIGHT <- 'Late Night'
NIGHT <- 'Night'
MORNING <- 'Morning'
AFTERNOON <- 'Afternoon'
EVENING <- 'Evening'

INTERVAL_5_6 <- '5 to 6'
INTERVAL_6_7 <- '6 to 7'
INTERVAL_7_8 <- '7 to 8'
INTERVAL_8_9 <- '8 to 9'
INTERVAL_9_10 <- '9 to 10'
INTERVAL_10_11 <- '10 to 11'
INTERVAL_11_12 <- '11 to 12'

accidents$Hour <- strftime(accidents$Start_Time, format="%H")

get_time_of_day <- function(hour) {
  if (hour >= 5 && hour < 12) {
    return(MORNING)
  } else if (hour >= 12 && hour < 16) {
    return(AFTERNOON)
  } else if (hour >= 16 && hour < 20) {
    return(EVENING)
  } else {
    return(NIGHT)
  }
}

accidents$Hour <- as.numeric(accidents$Hour)
accidents$time_of_the_day <- mapply(get_time_of_day, accidents$Hour)

Time_of_the_day <- accidents$time_of_the_day
ID <- accidents$ID
Severity <- accidents$Severity

bb <- aggregate(ID ~ Time_of_the_day, accidents, FUN=length)
bb$total = nrow(accidents)
bb <- xtabs(bb$ID / bb$total ~ bb$Time_of_the_day, bb)
barplot(t(bb)
        ,main="Comparison of proportion of accidents \noccuring at different times of the day"
        ,ylab="Proportion"
        ,xlab="Time of the day"
        ,col=c("mistyrose")
        ,las=1        
        ,args.legend=list(x="topleft")
        ,axes=F
        ,cex.axis=1.2
        ,cex.names=1.2
        ,cex.lab=1
        ,cex.main=1.2)
axis(side=2,at=c(0,0.1,0.2,0.3,0.4),labels=c("0","0.1","0.2","0.3","0.4"),las=2)
```

<b>The bar graph above illustrates the distribution of accidents across different times of the day, highlighting that the majority of accidents occur during daylight hours. Morning, afternoon, and evening collectively account for around 68% of all accidents, while the nighttime period represents approximately 30% of the total accidents.</b>

## Bar graph to display the hourly distribution of accidents

```{r}
accidents_new <- accidents %>% mutate(startHr=hour(Start_Time))
accidents_severity <- accidents_new %>% group_by(startHr) %>% summarise(mean(Severity))
accidents_count <- accidents_new %>%group_by(startHr, Severity) %>%summarise(n =n())
accident_summary <- merge(accidents_severity,accidents_count)
accident_summary <- accident_summary %>% rename(sev_avg = "mean(Severity)")
ggplot(data = accident_summary) + 
  geom_col(mapping=aes(x=startHr, y=n, fill=sev_avg)) +
  scale_fill_distiller(palette="Reds", trans= "reverse") +
  labs(
    title = "Amount of car accidents by hour",
    x = "Hour",
    y = "Number of accidents",
    caption = "A Countrywide Traffic Accident Dataset, 2016-2023.",
    fill = "Average Severity") +
  scale_y_continuous(expand = expansion(mult = c(0, .1)))
```

<b>The bar graph above visually displays the hourly distribution of accidents, with the legend indicating the severity caused by these accidents. Notably, the data reveals that the majority of accidents, approximately 60%, transpire between 5 AM and 5 PM. Subsequently, there is a gradual decrease in accident frequency from 5 PM until 3 AM the following morning. Interestingly, when examining accident severity, those transpiring between 6 PM and 10 PM tend to result in higher severity accidents.</b>

## Distribution of Accidents by Day & Hour

```{r}
accidents$hr <- lubridate::hour(accidents$Start_Time)
accidents$day <- lubridate::day(accidents$Start_Time)
accidents$week <- lubridate::week(accidents$Start_Time)
accidents$month <- lubridate::month(accidents$Start_Time)
accidents$Weekday <- lubridate::wday(accidents$Start_Time, label=T, abbr=F)
accidents$Weekdaya <- lubridate::wday(accidents$Start_Time, label=T, abbr=T)
accidents$Year <- lubridate::year(accidents$Start_Time)
accidents %>% group_by(hr, Weekday) %>%
  summarise(n = n()) %>%
  ggplot(aes(hr, n, fill = Weekday)) + 
  geom_bar(stat = "identity", show.legend = F) +
  facet_grid(~Weekday) +   labs(x = "24 hours of the day", y = "Count of Accidents", 
                                title = "Accidents by Day & Hour") + theme_light(base_size = 11) +
  scale_fill_brewer(palette=("Oranges")) + 
  theme(strip.background = element_rect(fill="grey90")) +
  theme(strip.text = element_text(size = 13, color = 'black')) +
  scale_y_continuous(labels = scales::comma)
```

<b>The graph depicted above illustrates the distribution of accidents across various days. It is evident that the majority of accidents have occurred on weekdays, with Thursdays and Fridays leading the statistics, accounting for a combined total of 34.77% of all accidents. Wednesdays follow closely, making up around 17% of the total accidents. Tuesdays account for 16.6%, Mondays for 15.6%, while Saturdays and Sundays collectively represent 16% of the total accidents.</b>

## Distribution of accidents in various weather conditions

```{r}
accidents %>%
  group_by(Weather_Condition) %>%
  filter(Severity==4) %>%
  summarise(count = n()/nrow(.)*100) %>%
  filter(count>1)%>%
  arrange(-count) -> severity4_weather
severity4_weather

#options(repr.plot.width = 20, repr.plot.height = 8)
severity4_weather %>%
  ggplot() +
  geom_col(mapping=aes(x=reorder(Weather_Condition, -count), y=count, fill=Weather_Condition))+ 
  theme(text = element_text(size=10),plot.title = element_text(hjust = 0.5, vjust = 1),legend.key.width = unit(0.5, "cm"), legend.key.height = unit(0.25, "cm")) +
  labs(x = "Weather condition", y="%", title ="Severity 4 acccidents by weather")-> severity4_weather_plot

accidents %>%
  group_by(Weather_Condition) %>%
  filter(Severity==3) %>%
  summarise(count = n()/nrow(.)*100) %>%
  filter(count>1)%>%
  arrange(-count) -> severity3_weather
severity3_weather

#options(repr.plot.width = 20, repr.plot.height = 8)
severity3_weather %>%
  ggplot() +
  geom_col(mapping=aes(x=reorder(Weather_Condition, -count), y=count, fill=Weather_Condition))+ 
  theme(text = element_text(size=10),plot.title = element_text(hjust = 0.5, vjust = 1),legend.key.width = unit(0.5, "cm"), legend.key.height = unit(0.25, "cm")) +
  labs(x = "Weather condition", y="%", title ="Severity 3 acccidents by weather") -> severity3_weather_plot

accidents %>%
  group_by(Weather_Condition) %>%
  filter(Severity==2) %>%
  summarise(count = n()/nrow(.)*100) %>%
  filter(count>1)%>%
  arrange(-count) -> severity2_weather
severity2_weather

#options(repr.plot.width = 20, repr.plot.height = 8)
severity2_weather %>%
  ggplot() +
  geom_col(mapping=aes(x=reorder(Weather_Condition, -count), y=count, fill=Weather_Condition))+ 
  theme(text = element_text(size=10),plot.title = element_text(hjust = 0.5, vjust = 1),legend.key.width = unit(0.5, "cm"), legend.key.height = unit(0.25, "cm")) +
  labs(x = "Weather condition", y="%", title ="Severity 2 acccidents by weather") -> severity2_weather_plot

accidents %>%
  group_by(Weather_Condition) %>%
  filter(Severity==1) %>%
  summarise(count = n()/nrow(.)*100) %>%
  filter(count>1)%>%
  arrange(-count) -> severity1_weather
severity1_weather

severity1_weather %>%
  ggplot() +
  geom_col(mapping=aes(x=reorder(Weather_Condition, -count), y=count, fill=Weather_Condition))+ 
  theme(text = element_text(size=10),plot.title = element_text(hjust = 0.5, vjust = 1),legend.key.width = unit(0.5, "cm"), legend.key.height = unit(0.25, "cm"))+
  labs(x = "Weather condition", y="%", title ="Severity 1 acccidents by weather")-> severity1_weather_plot


combined_plot <- (severity4_weather_plot | severity3_weather_plot) / (severity2_weather_plot | severity1_weather_plot)

# Set the size for the combined plot
options(repr.plot.width = 120, repr.plot.height = 12)

# Display the combined plot
combined_plot
```


<b>The graphs above depict the distribution of accidents in various weather conditions for each severity level. It is discernible that the majority of accidents have taken place when the weather was either fair or cloudy, regardless of severity level. In essence, nearly 60-70% of accidents resulting in severity levels 1, 2, or 3 occurred under fair or cloudy weather conditions, and a substantial 97% of severity 4 accidents also transpired under fair or cloudy weather conditions. The remaining 3% of severity 4 accidents were associated with light rain, with no other weather conditions significantly impacting accidents leading to Severity 4.</b>



# Hypothesis Testing
**Hypothesis: There is a difference in the average number of accidents across different months of the year.
•	Null Hypothesis (H0): The average number of accidents is the same for all months.
•	Alternative Hypothesis (H1): The average number of accidents varies by month.**
```{r}
accidents$month <- lubridate::month(accidents$Start_Time)
accidents$accident_count <- 1

accidents$month = factor(accidents$month, order=T)
anova_result <- aov(accident_count ~ month, data = accidents)
summary(anova_result)

xkabledply(anova_result, title = "ANOVA result summary") 
```

The analysis does not provide enough evidence to conclude that there is a statistically significant difference in the average number of accidents across different months of the year. Therefore, we fail to reject the null hypothesis, and the data does not support the idea that the month of the year has a significant impact on the number of accidents.

**Hypothesis: Accidents are more likely to occur on weekends compared to weekdays.
•	Null Hypothesis (H0): The probability of an accident on weekends is the same as on weekdays.
•	Alternative Hypothesis (H1): The probability of an accident on weekdays is greater than on weekends.**
```{r}
accidents$day_type <- ifelse(weekdays(accidents$Start_Time) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")

contingency_table <- table(accidents$day_type)
print(contingency_table)
chi_square_result <- chisq.test(contingency_table, correct = FALSE)
#xkabledply(chi_square_result)
print(chi_square_result)
```

The results of the chi-squared test provide strong evidence to reject the null hypothesis. The data supports the alternative hypothesis, which implies that the probability of accidents is greater on weekdays than on weekends.

**Hypothesis: The time of day (morning, afternoon, evening, night) has an effect on the occurrence of fatal accidents.
•	Null Hypothesis (H0): The time of day does not affect the occurrence of fatal accidents.
•	Alternative Hypothesis (H1): The time of day affects the occurrence of fatal accidents.**

```{r}
LATE_NIGHT <- 'Late Night'
NIGHT <- 'Night'
MORNING <- 'Morning'
AFTERNOON <- 'Afternoon'
EVENING <- 'Evening'

accidents$Hour <- lubridate::hour(accidents$Start_Time)
get_time_of_day <- function(hour) {
  if (hour >= 5 && hour < 12) {
    return(MORNING)
  } else if (hour >= 12 && hour < 16) {
    return(AFTERNOON)
  } else if (hour >= 16 && hour < 20) {
    return(EVENING)
  } else {
    return(NIGHT)
  }
}

accidents$time_of_the_day <- mapply(get_time_of_day, accidents$Hour)

contingency_table <- table(accidents$time_of_the_day)
print(contingency_table)
chi_square_result <- chisq.test(contingency_table, correct = FALSE)
print(chi_square_result)
```

The results of the chi-squared test provide strong evidence to reject the null hypothesis. The data supports the alternative hypothesis, indicating that the time of day does affect the occurrence of fatal accidents. This suggests that there is a statistically significant relationship between the time of day and the occurrence of fatal accidents.
